An Interpretable Machine Learning Framework for Modeling High-Resolution Spectroscopic Data\footnote{Open source code at \url{https://github.com/gully/blase}}

### 2022-08-15 paper/paper1/ms.tex
Shorten the abstract to meet 250 words, on CCA rooftop :)

    Comparison of \'echelle spectra to synthetic models has become a computational statistics challenge, with over ten thousand individual spectral lines affecting a typical cool star \'echelle spectrum.  Telluric artifacts, imperfect line lists, inexact continuum placement, and inflexible models frustrate the scientific promise of these information-rich datasets.  Here we debut an interpretable machine-learning framework \emph{blas\'e} that addresses these and other challenges.  The semi-empirical approach can be viewed as ``transfer learning''--first pre-training models on noise-free precomputed synthetic spectral models, then learning the corrections to line depths and widths from whole-spectrum fitting to an observed spectrum.  The auto-differentiable model employs back-propagation, the fundamental algorithm empowering modern Deep Learning and Neural Networks. Here, however, the 40,000+ parameters symbolize physically interpretable line profile properties such as amplitude, width, location, and shape, plus radial velocity and rotational broadening.  This hybrid data-/model- driven framework allows joint modeling of stellar and telluric lines simultaneously, a potentially transformative step forwards for mitigating the deleterious telluric contamination in the near-infrared.  The \emph{blas\'e} approach also acts as a deconvolution tool. The general purpose scaffolding may be extensible to many scientific applications, including Precision Radial Velocities, Doppler Imaging, chemical abundances for Galactic archaeology, line veiling, magnetic fields, and remote sensing.  Blas\'e can create super-resolution semi-empirical templates useful for critically evaluating imperfections in atomic and molecular line lists.  Its sparse-matrix architecture and GPU-acceleration make \emph{blas\'e} fast, with end-to-end training of an octave of bandwidth (50+ \'echelle orders) in under 1 minute.  The open-source PyTorch-based code \texttt{blase} includes tutorials, deep dives, reproducible Python environments, and API documentation.  We show how the tool fits into the existing Python spectroscopy ecosystem, demonstrate a range of astrophysical applications, and discuss limitations and future extensions.

>>>>

    Comparison of \'echelle spectra to synthetic models has become a computational statistics challenge, with over ten thousand individual spectral lines affecting a typical cool star \'echelle spectrum.  Telluric artifacts, imperfect line lists, inexact continuum placement, and inflexible models frustrate the scientific promise of these information-rich datasets.  Here we debut an interpretable machine-learning framework \emph{blas\'e} that addresses these and other challenges.  The semi-empirical approach can be viewed as ``transfer learning''--first pre-training models on noise-free precomputed synthetic spectral models, then learning the corrections to line depths and widths from whole-spectrum fitting to an observed spectrum.  The auto-differentiable model employs back-propagation, the fundamental algorithm empowering modern Deep Learning and Neural Networks. Here, however, the 40,000+ parameters symbolize physically interpretable line profile properties such as amplitude, width, location, and shape, plus radial velocity and rotational broadening.  This hybrid data-/model- driven framework allows joint modeling of stellar and telluric lines simultaneously, a potentially transformative step forwards for mitigating the deleterious telluric contamination in the near-infrared.  The \emph{blas\'e} approach acts as both a deconvolution tool and semi-empirical model. The general purpose scaffolding may be extensible to many scientific applications, including Precision Radial Velocities, Doppler Imaging, chemical abundances for Galactic archaeology, line veiling, magnetic fields, and remote sensing.  Its sparse-matrix architecture and GPU-acceleration make \emph{blas\'e} fast.  The open-source PyTorch-based code \texttt{blase} includes tutorials, Application Programming Interface (API) documentation, and more.  We show how the tool fits into the existing Python spectroscopy ecosystem, demonstrate a range of astrophysical applications, and discuss limitations and future extensions.
